---
title: "WORKING TITLE"
subtitle: "Week 2"
author: "Kyle Nessen"
date: "2024-01-17"
toc: true
categories: []
format: 
   html:
     df-print: kable
execute:
  echo: false
---

```{r}
#| include: false
library(tidyverse)
library(RMySQL)
con <- dbConnect(RMySQL::MySQL(), dbname = "selectree", host = "localhost", 
                 port = 3306, user = "rstudio", password = "rstudio321")
```

## Previous week meeting notes

-   Matt, May, and I met ahead of our meeting with Daniel and Logan.

-   Matt identified that missing sunset species likely were not in The New Western Garden Book, and therefore there is nothing to look up in the book or on other websites.

-   Matt told me to give him an equivalent species column for him to fill out which we can draw sunset zones from.

-   May had me remove the zone columns in the Google Spreadsheet, as it would be easier to copy and paste from the SelecTree website.

-   I offered to try to do a look-up with code for the missing sunset zones first before getting Caitlin involved.

-   In terms of taxonomy, Matt wants to be presented a list of scientific names that are currently not accepted by a name parsing authority. He will go through and decide if it is a true synonym or not.

-   Additionally, there is no need for full authorship names. Goal is to make it easy for people to find the tree species they are looking for.

-   Meeting with Daniel and Logan was helpful. They provided their internal documentation for how the database is organized

    -   <https://muchquak.github.io/database/tables/>

-   Ideal format to send changes to them is a series of SQL statements that will update specific rows of the database. Alternatively csv files will work.

    -   More discussion is needed as both approaches sound messy and cumbersome.

-   Lots of discussion about how I can access the database. Ultimately we settled on a database dump, which is a local copy I can explore. Logan sent this shortly after the meeting.

## Database access

I'm very pleased to get access to the website database. The data is much more organized and it is far easier to find issues and incomplete information.

For example, last week I identified a number of taxa that were missing the specific epithet in the scientific name. In order to find them, I had to pass the names through GBIF, and filter by those that had poor scores. This is an overly complicated process that is likely error prone.

Now, I can simply pull up the `taxon` table and see *exactly* which names are missing in the `species` column. Here are the first ten as an example of what the taxon table looks like:

```{r}
taxon <- dbReadTable(con, "taxon")
taxon %>% filter(species == "") %>% head(10)
```

Last week, I said there was at least `43` incomplete names, but the actual number is `59`. Some of those new names are synonyms, however.

Here is a new Google Sheet with those species. This seems like a good task for **Caitlin** to fill out the missing info. I've highlighted the column that needs to be completed, but she should use additional columns as needed.

```{r}
missing_names <- taxon %>% filter(species == "")
write_csv(missing_names,"missing_scientific_names_20240122.csv")
```

```{=html}
<div style="text-align: center;">
    <a href="https://docs.google.com/spreadsheets/d/1M8G44zQhIH3dLPVRxgsRYe7laZQMAnzr5vBvO-duObk/edit?usp=sharing" target="_blank" style="background-color: #4CAF50; /* Green */
        border: none;
        color: white;
        padding: 20px 40px; /* Size of the button */
        text-align: center;
        text-decoration: none;
        display: inline-block;
        font-size: 16px; /* Font size */
        margin: 4px 2px;
        cursor: pointer;
        border-radius: 12px;">SelecTree - Missing Specific Epithets</a>
</div>
```
## Sunset Zones

After our meeting last week, Matt went through and added analogous tree species for our missing sunset zones. I took those names and looked up the zone values we have stored already.

Only two species had issues:

1.  *Ulmus chinensis* is not in SelecTree. I assume Matt meant *Ulmus parvifolia*. Let me know if that's not the case.
2.  *Vitex negundo* is not in SelecTree (only the cultivar), and I cannot find anything online for sunset zones. **This is the only missing tree for sunset zones**. If you give me an alternative, I can easily fix it.

I've generated a csv below that should be straightforward to incorporate into the production database of the website. I can go into the details in our meeting if you want, but essentially each row represents a sunset zone, with tree_id repeating if a species has multiple zones.

This new csv is almost 1000 rows long, so writing out SQL statements as Logan suggested is impractical. A bit of searching indicates that he should be able to load this csv into some scripting language and insert the new values easily. I could also do this, but I don't have access to their SQL database and I'm still not an admin on the website. I think long-term, everyone would be happier if we reduce middlemen, so I would like to explore options on how I can update the database directly.

Let me know if you'd like me to coordinate with Logan about the updated sunset zones csv.

```{r}
#| include: false
library(readr)
matt_zones <- read_csv("missing_sunset_zones_matt_completed_20240122.csv")

# Perform the join
matt_zones <- matt_zones %>%
  left_join(taxon, by = c("Analgous Tree Species" = "name_unformatted"))

# Manually add Osmanthus × fortunei
matt_zones$tree_id.y[23] <- 977

# TODO add value for Vitex negundo 'Heterophylla'
#matt_zones$tree_id.y[72] <- 

matt_zones <- matt_zones %>%
  select(tree_id.x, tree_id.y) %>%
  rename(
    tree_id = tree_id.x,
    reference_tree_id = tree_id.y
  ) %>%
  filter(complete.cases(.)) # Will filter out Vitex

sunset_zones <- dbReadTable(con, "sunset_zone")

zones_joined <- matt_zones %>%
  left_join(sunset_zones, by = c("reference_tree_id" = "tree_id"))

zones_joined <- zones_joined %>% select(c(tree_id,sunset_zone))

write_csv(zones_joined,"db_ready_sunset_zones_20240122.csv")

```

::: {style="text-align: center;"}
```{r}
library(downloadthis)
zones_joined %>%
  download_this(
    output_name = "db_ready_sunset_zones",
    output_extension = ".csv",
    button_label = "Download updated sunset zones as csv",
    button_type = "default",
    has_icon = TRUE,
    icon = "fa fa-save",
    csv2 = FALSE
  )
```
:::


